---
title: "Additive Genetic Models in Mixed Populations"
author: "Facundo Muñoz"
date: '`r paste(Sys.Date(), 'breedR version:', packageVersion('breedR'))`'
output:
  pdf_document:
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    number_sections: yes
    toc: yes
    toc_depth: 1
  md_document:
    toc: yes
    toc_depth: 1
    variant: markdown_github
vignette: >
  %\VignetteIndexEntry{Additive Genetic Models in Mixed Populations}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
  \usepackage{amsmath}
---

```{r setup-knitr, include=FALSE, purl=FALSE, eval = FALSE}
library(knitr)
opts_chunk$set(echo       = TRUE,
               message    = TRUE,
               warning    = TRUE,
#                comment    = NA,
               fig.width  = 4,
               fig.height = 3,
               cache      = TRUE)
```

```{r setup-parallel-computing}
library(doParallel)
cl <- makeCluster(8)
registerDoParallel(cl)
```


```{r functions}

## Given named vectors with the number of individuals by population
## and their corresponding population variances, return a dataframe
## with and id, the population index and a simulated Breeding Value
sample_founders <- function(n.founders, sigma2) {
  
  stopifnot(identical(names(n.founders), names(sigma2)))
  nms <- names(n.founders)
  
  cfapply <- function(fun) do.call('c', lapply(nms, fun))
  
  founders <- data.frame(id = cfapply(function(x) paste0(x, 1:n.founders[x])),
                         pop.idx = cfapply(function(x) rep(match(x, nms), n.founders[x])),
                         BV = cfapply(function(x) rnorm(n.founders[x], sd = sqrt(sigma2[x]))))
  return(founders)
}

sample_dat <- function (n.founders, sigma2, n.obs) {

  ## The 'family' is independent of the order of the parents
  ## While the population only takes into account the origin
  ## e.g. cross2fam(c('J3', 'E1')) gives 'E1:J3'
  ## while cross2pop(c('J3', 'E1')) gives 'EJ'
  cross2fam <- function(x) paste(founders$id[sort(x)], collapse = ':')
  cross2pop <- function(x) paste(names(n.founders)[founders$pop.idx[sort(x)]], collapse = '')
  
  
  founders <- sample_founders(n.founders, sigma2[names(n.founders)])
  
  ## A balanced design with no self-crossings
  ## The observed problems are not due to diagonals, though.
  #   n.reps <- 20
  #   crossings <- which(lower.tri(matrix(TRUE, nrow(founders), nrow(founders))), arr.ind = TRUE, useNames = FALSE)
  #   obs.parents.idx <- crossings[rep(1:nrow(crossings), each = n.reps), ]
  #   n.obs <- nrow(obs.parents.idx)
  
  ## Simulate crossings (random)
  # n.obs <- sum(n.founders)*150
  obs.parents.idx <- matrix(sample(nrow(founders), 2*n.obs, replace = TRUE), ncol = 2)
  
  dat <- data.frame(
    id  = sum(n.founders) + seq.int(n.obs),
    dad = obs.parents.idx[, 1],
    mum = obs.parents.idx[, 2],
    fam = apply(obs.parents.idx, 1, cross2fam),
    sp  = apply(obs.parents.idx, 1, cross2pop),
    bv  = apply(obs.parents.idx, 1, 
                function(x) mean(founders$BV[x]) + msp(x)),
    resid = rnorm(n.obs, sd = sqrt(sigma2['resid'])))
  
  dat <- transform(dat,
                   y = bv + resid)
  return(dat)
}

## Mendelian sampling term
## For self-polinated crossings, this should be smaller.
msp <- function(x) {
  ss <- sigma2[founders$pop.idx[x]]
  s2 <- (ss[1] + ss[2])/4
  rnorm(1, sd = sqrt(s2))
}

## Give the index vector of additive-genetic random effects
## that belong to one subpopulation;
## 'E', 'J' (founders) or 'EE', 'EJ' or 'JJ' (offspring).
idx_pop <- function(x) {
  if (nchar(x) == 1) grep(x, founders$id)
  else
    match(dat$id[dat$sp == x], as.data.frame(ped)$self)
}
```

Full diallel trial with founders from two different base populations E and J

```{r diallel-setup, warning = FALSE}
## Setup
library(breedR)
library(ggplot2)
set.seed(1)

## Simulation parameters
n.founders <- c(E = 9, J = 9)
sigma2 <- c(E = 9, J = 6, resid = 1)
n.obs  <- 3000

founders <- sample_founders(n.founders, sigma2[names(n.founders)])
dat <- sample_dat(n.founders, sigma2, n.obs)

## Check: variability of the realised blups of hybrids
# sv <- sapply(1:1e2, function(x) {dat <- sample_dat(); var(dat$bv[dat$sp=='EJ'])})
# hist(sv)

## Printing simulated setting
print(table(dat[, c('mum', 'dad')]), zero.print = "")
str(dat)
```


There will be two independent additive-genetic variance models for the  J  and  E  populations.
Each variance component will be estimated using the *pure* offspring only.

```{r overall-genetic-structure}

## Build a pedigree for the whole mixed population
## and get the kinship matrix A
ped <- build_pedigree(1:3, data = dat)
A <- pedigreemm::getA(ped)

## Build the full incidence matrix
Z <- as(dat$id, 'indMatrix')
```


## Method 1: Hybrids as an independent population

This is the easiest way to go.
It works as a first approximation, but has several shortcommings.

It needs to estimate one *virtual* variance for the hybrid population which is not linked to any genetic variance in the *real world*.
Moreover, we don't use the hybrid observations to learn about the two varainces that really matter: $\sigma 2_E$ and $\sigma 2_J$

The advantage is that it predicts the Breeding Values of the hybrid offspring.
However, the accuracy may be limited by the violation of the *single population* hypothesis of the model.

```{r fit1}
## Avoid estimating BLUPS for which we don't have information
## Otherwise, the run takes much longer (5 hs vs 6 min in this example)


## A[idx_pop('EE'), idx_pop('JJ')]  # This is null: populations are independent
Z_EE <- Z[, idx_pop('EE')]
Z_EJ <- Z[, idx_pop('EJ')]
Z_JJ <- Z[, idx_pop('JJ')]

A_EE <- A[idx_pop('EE'), idx_pop('EE')]
A_EJ <- A[idx_pop('EJ'), idx_pop('EJ')]
A_JJ <- A[idx_pop('JJ'), idx_pop('JJ')]

## Now fit a model with three additive-genetic compnents, 
## by means of generic effects (as only one 'genetic' is allowed in breedR)

res1 <- remlf90(y ~ sp,
                generic = list(
                  E = list(incidence  = Z_EE,
                           covariance = A_EE),
                  J = list(incidence  = Z_JJ,
                           covariance = A_JJ),
                  H = list(incidence  = Z_EJ,
                           covariance = A_EJ)),
                data = dat
)
```


```{r fit1-summary}
summary(res1)
```


```{r fit1-predicted-breeding-values}
PBV <- as.matrix(cbind(Z_EE, Z_JJ, Z_EJ)) %*%
  do.call('rbind', lapply(ranef(res1), function(x) cbind(PBV = x, se = attr(x, 'se'))))

ggplot(cbind(dat, PBV), aes(bv, PBV)) +
  geom_point() +
  geom_abline(int = 0, sl = 1, col = 'darkgray')
```


## Method 2: GCA/SCA model for hybrids

Another approach is to model hybrids with GCAs and SCAs, with additional variance parameters.
The additive component of the genetic variances for the E and J populations is half the population variance.
But these parameters will also gather dominance and epistasis effects.

However, we don't take advantage of the known relationship between the additive components by treating them as independent parameters.
Furthermore, we don't use the relationship between hybrids and pures to learn about the original genetic variances.
Nor even within hybrids, as we treat SCA as an unstructured effect, while there are both half and full siblings.



```{r fit2}
## We only want to apply 'dad', 'mum' and 'sca' effects to hybrids,
## and make it zero for non-hybrids. We do so by pre-multiplying by a 
## diagonal indicator matrix
Ind <- diag(dat$sp == 'EJ')

## Build a specific incidence matrices for generic random effects
Z_dad <- Ind %*% as(dat$dad, 'indMatrix')
Z_mum <- Ind %*% as(dat$mum, 'indMatrix')
Z_sca <- Ind %*% as(as.numeric(dat$fam), 'indMatrix')

## The structure variances are diagonal
D <- diag(sum(n.founders))

res2 <- remlf90(y ~ sp,
                generic = list(
                  E = list(incidence  = Z_EE,
                           covariance = A_EE),
                  J = list(incidence  = Z_JJ,
                           covariance = A_JJ),
                  dad = list(incidence = Z_dad,
                             covariance = D),
                  mum = list(incidence = Z_mum,
                             covariance = D),
                  sca = list(incidence = Z_sca,
                             covariance = diag(nlevels(dat$fam)))),
                data = transform(dat)
)
```

```{r fit2-summary}
summary(res2)
```


```{r fit2-predicted-breeding-values}
PBV <- as.matrix(cbind(Z_EE, Z_JJ, Z_dad, Z_mum, Z_sca)) %*%
  do.call('rbind', lapply(ranef(res2), function(x) cbind(PBV = x, se = attr(x, 'se'))))

ggplot(cbind(dat, PBV), aes(bv, PBV)) +
  geom_point() +
  geom_abline(int = 0, sl = 1, col = 'darkgray')
```


This approach does not work very well.
Much of the variation has not been adequately accounted for, and ended up in the residuals.
May be there is some misspecification in the model.


## Method 3: grid search

The difficulty of the problem resides that we cannot estimate two variance parameters at the same time.
If we split the matrix $A$ in blocks corresponding to each subpopulation, we can write the covariance matrix for the mixture as
\begin{equation}
  \begin{split}
    \Sigma & =
    \begin{bmatrix}
      \sigma^2_E A_{E}    & \mathbf{0}        & \sigma^2_E A_{E:EE} & \frac{3\sigma^2_E + \sigma^2_J}{4} A_{E:H} & \mathbf{0} \\
      \mathbf{0}          & \sigma^2_J A_{J}  & \mathbf{0} & \frac{\sigma^2_E + 3\sigma^2_J}{4} A_{J:H} & \sigma^2_J A_{J:JJ} \\
      \sigma^2_E A_{E:EE} & \mathbf{0}        & \sigma^2_E A_{EE} & \frac{3\sigma^2_E + \sigma^2_J}{4} A_{EH} & \mathbf{0} \\
      \frac{3\sigma^2_E + \sigma^2_J}{4} A_{E:H} & \frac{\sigma^2_E + 3\sigma^2_J}{4} A_{J:H} & \frac{3\sigma^2_E + \sigma^2_J}{4} A_{HE} & \frac{\sigma^2_E + \sigma^2_J}{2} A_{HH} & \frac{\sigma^2_E + 3\sigma^2_J}{4} A_{HJ} \\
      \mathbf{0}          & \sigma^2_J A_{J:JJ} & \mathbf{0} & \frac{\sigma^2_E + 3\sigma^2_J}{4} A_{JH} & \sigma^2_J A_{JJ} \\
    \end{bmatrix} \\
    & = \sigma^2_E
    \begin{bmatrix}
       A_{E}      & \mathbf{0}  &  A_{E:EE}   & \frac{3 + \lambda}{4} A_{E:H}  & \mathbf{0} \\
      \mathbf{0}  &  \lambda A_{J}      & \mathbf{0}  & \frac{1 + 3\lambda}{4} A_{J:H}  & \lambda A_{J:JJ} \\
       A_{E:EE} & \mathbf{0}        & A_{EE} & \frac{3 + \lambda}{4} A_{EH} & \mathbf{0} \\
      \frac{3 + \lambda}{4} A_{E:H} & \frac{1 + 3\lambda}{4} A_{J:H} & \frac{3 + \lambda}{4} A_{HE} & \frac{1 + \lambda}{2} A_{HH} & \frac{1 + 3\lambda}{4} A_{HJ} \\
      \mathbf{0}          & \lambda A_{J:JJ} & \mathbf{0} & \frac{1 + 3\lambda}{4} A_{JH} & \lambda A_{JJ} \\
    \end{bmatrix},
  \end{split}
\end{equation}
where $\lambda = \frac{\sigma^2_J}{\sigma^2_E}$.

However, if we are not interested in evaluating the parents, we can disregard the first two block rows and columns from the matrix.

Now, fit the model for several values of $\lambda$ and maximize the likelihood.

```{r likelihood-profiling}

## Introduce the corresponding scaling factors 
## in the relationship matrix
scale_A <- function(x) {

  A <- pedigreemm::getA(ped)
  
  ## The pure E subpopulations remains the same
  S <- A
  E.idx <- c(idx_pop('E'), idx_pop('EE'))
  
  ## The pure J subpopulations get multiplied by lambda
  J.idx <- c(idx_pop('J'), idx_pop('JJ'))
  S[J.idx, J.idx] <- A[J.idx, J.idx] * x
  
  ## The hybrids related wuth pure E get a factor of (3+lambda)/4
  S[idx_pop('EJ'), E.idx] <- A[idx_pop('EJ'), E.idx] * (3+x)/4
  S[E.idx, idx_pop('EJ')] <- A[E.idx, idx_pop('EJ')] * (3+x)/4
  
  ## The hybrids related wuth pure J get a factor of (1+3*lambda)/4
  S[idx_pop('EJ'), J.idx] <- A[idx_pop('EJ'), J.idx] * (1+3*x)/4
  S[J.idx, idx_pop('EJ')] <- A[J.idx, idx_pop('EJ')] * (1+3*x)/4

  ## Finally, the hybrids related with other hybrids get a factor of (1+lambda)/2
  S[idx_pop('EJ'), idx_pop('EJ')] <- A[idx_pop('EJ'), idx_pop('EJ')] * (1+x)/2

  return(S)
}

## Condicional likelihood given lambda
cond_lik <- function(x) {
  require(breedR)
  ## Conditional structure matrix
  S <- scale_A(x)
  
  #   ## Temporarily, let's use only the pure pops
  #   idx <- c(idx_pop('EE'), idx_pop('JJ'))
  idx <- -c(idx_pop('E'), idx_pop('J'))  # exclude founders
  
  suppressWarnings(
    res <- remlf90(y ~ sp,
                   generic = list(
                     E = list(incidence  = Z[, idx],
                              covariance = S[idx, idx])),
                   data = dat
    )
  )
  logLik(res)
}

lambda <- seq(.3, 1, length.out = 11)

# lik <- sapply(lambda, cond_lik)  # (sequential)
lik <- foreach(x = seq.int(lambda), .combine = c) %dopar% cond_lik(lambda[x])

ggplot(data.frame(lambda, lik)[6:11,], aes(lambda, lik)) + 
  geom_line()
```

This one works well.
But for some reason, when I include the hybrids, for lambdas below about 0.58 all the variance is accounted for residual variance.

There must be something wrong somewhere. May be the simulation of the Breeding Values is not correct?

Analogously, if I include only the hybrid subpopulation, it also works quite well, identifying the base variance.


```{r fit3}

## Take lambda maximizing the likelihood
lambda0 <- lambda[which.max(lik)]

S <- scale_A(lambda0)

## Temporarily, let's use only the pure pops
idx <- c(idx_pop('EE'), idx_pop('JJ'))

# ## Remove the founders, which I don't want to evaluate
# idx <- -(1:sum(n.founders))

res <- remlf90(y ~ sp,
               generic = list(
                 E = list(incidence  = Z[dat$sp != 'EJ', idx],
                          covariance = S[idx, idx])),
               data = dat[dat$sp != 'EJ', ]
)

res3 <- remlf90(y ~ 1,
                generic = list(
                  E = list(incidence  = Z[,-(1:18)],
                           covariance = S[-(1:18), -(1:18)],
                           var.ini = 3)),
                var.ini = list(resid = 1),
                data = dat)

```


```{r fit3-summary}
summary(res3)
```

Lambda was maximized at `r lambda0`, giving an estimated additive-genetic variance for the `J` population of `r round(res3$var['E', 1]*lambda0,2)`.

```{r fit3-predicted-breeding-values}
PBV <- as.matrix(Z[dat$sp != 'EJ', idx]) %*%
  do.call('rbind', lapply(ranef(res3), function(x) cbind(PBV = x, se = attr(x, 'se'))))

ggplot(cbind(dat[dat$sp != 'EJ', ], PBV), aes(bv, PBV)) +
  geom_point() +
  geom_abline(int = 0, sl = 1, col = 'darkgray')
```



## Trying to identify the problem

For low values of $\lambda$, the model favours the residual component rather than the structured random effect.
I believe this might be due to some misspecification of the model, or error in the simulation of data.

Hoever, the following results show that the simulated data are compatible with the assumed model.

It is not a matter of the initial variance. 
I have tried specifying the true variances as initial values, and the problem persists.

It is not a matter of the null fixed effects.
I have tried specifying only an intercept, and the problem persists.

It is not a matter of variance magnitudes.
I have tried specifying variances of 9, 6 and 1 for E, J and residuals respectively, and the problem persists.

It is not a matter of unbalanced sampling.
I have designed a balanced simulation, and the problem persists.

It is not a problem of misspecification of the Mendelian Sampling Term for self-crossings.
In effect, it is not exactly how we defined it. But I excluded the self-cronssings from the desing, and the problem persists.

It is not a problem of including the founders in the effect.
I have tried using only the observed blocks of both the incidence and covariance matrix, and the problem persists.

```{r detecting-outliers}

## Which of the simulated BVs are most incompatible
## with the assumed covariance matrix S?

## Set each simulated BV to its marginal mean (0)
## and measure the Mahalanobis distance between the 
## modified vector of BVs to the actually observed.
## This equals the BV in question times the square root
## of the element in the corresponding position of the
## precision matrix.
## The BVs with higher distance to the mean are the 
## most uncompatible.

lambda0 <- 2/3

ped <- build_pedigree(1:3, data = dat)
A <- pedigreemm::getA(ped)
S <- scale_A(lambda0)

## Marginalize observed variables only, and scale
Sigma <- sigma2['E']*S[-(1:sum(n.founders)), -(1:sum(n.founders))]
Q <- solve(Sigma)  ## This is very unstable!! (A is very ill-conditioned)

md <- dat$bv * sqrt(diag(Q))

plot(md, pch = 20)
hist(md, prob = TRUE)
curve(dnorm(x, mean = 0, sd = sqrt(var(md))),
     add = TRUE)
qqnorm(md, pch = 20)
qqline(md)

ggplot(cbind(dat, md), aes(md, fill = sp)) + 
  geom_density(alpha = 0.4)
```



```{r comparing-likelihoods, eval = FALSE}
## This does not work well:
## Usually det(Q) equals 0, because of malconditioning
## and its logarithm goes to infinity.
dat <- sample_dat(n.founders, sigma2, n.obs = 100)

loglik <- function(x, sigmaE, lambda0, data = dat) {
  ped <- breedR::build_pedigree(1:3, data = dat)
  S <- sigmaE * scale_A(lambda0)[-(1:18), -(1:18)]
  Q <- solve(S)

  k <- length(x)

  x %*% Q %*% x - log(det(Q))/2 - k*log(2*pi)/2
}

lik <- foreach(lmb = lambda) %dopar% lambda
  foreach(s2e = 1:15) %dopar% loglik(dat$bv, s2e, lmb, dat)

sapply(lambda0,
       function(lmb) sapply(1:,
                            function(s2e)))
dat$bv %*% Q %*% dat$bv
dat$bv %*% diag(nrow(dat)) %*% dat$bv
```



Can be a problem of malconditioned covariance Matrix.
Already, A is quite ill-conditioned.
When scaling with $\lambda$ small, the condition gets worse.

May a solution be using the smaller variance as a reference, so that $\lambda$ is always greater than 1?
This does not help. It only reverses the problem so that the condition gets worse for higher $\lambda$.

In summary, it seems to be a problem of malconditioning, which worsens for ratios of variances $\lambda$ more extremes than the true ration.
Furthermore, it must have something to do with the relationship between *hybrids* and *pures*, for when these are excluded, everything works fine, even for extreme values of $\lambda$

```{r likelihood-profiling-the-other-way-around}

## Introduce the corresponding scaling factors 
## in the relationship matrix
revscale_A <- function(x) {

  A <- pedigreemm::getA(ped)
  
  ## The pure E subpopulations get multiplied by tau
  S <- A
  E.idx <- c(idx_pop('E'), idx_pop('EE'))
  S[E.idx, E.idx] <- A[E.idx, E.idx] * x
  
  ## The pure J subpopulations remains the same
  J.idx <- c(idx_pop('J'), idx_pop('JJ'))
  
  ## The hybrids related wuth pure E get a factor of (3*tau+1)/4
  S[idx_pop('EJ'), E.idx] <- A[idx_pop('EJ'), E.idx] * (3*x+1)/4
  S[E.idx, idx_pop('EJ')] <- A[E.idx, idx_pop('EJ')] * (3*x+1)/4
  
  ## The hybrids related wuth pure J get a factor of (tau+3)/4
  S[idx_pop('EJ'), J.idx] <- A[idx_pop('EJ'), J.idx] * (x+3)/4
  S[J.idx, idx_pop('EJ')] <- A[J.idx, idx_pop('EJ')] * (x+3)/4

  ## Finally, the hybrids related with other hybrids get a factor of (1+lambda)/2
  S[idx_pop('EJ'), idx_pop('EJ')] <- A[idx_pop('EJ'), idx_pop('EJ')] * (x+1)/2

  return(S)
}

## Condicional likelihood given lambda
revcond_lik <- function(x) {
  require(breedR)
  ## Conditional structure matrix
  S <- revscale_A(x)
  
  #   ## Temporarily, let's use only the pure pops
  #   idx <- c(idx_pop('EE'), idx_pop('JJ'))
  idx <- -c(idx_pop('E'), idx_pop('J'))  # exclude founders
  
  suppressWarnings(
    res <- remlf90(y ~ sp,
                   generic = list(
                     J = list(incidence  = Z[, idx],
                              covariance = S[idx, idx])),
                   data = dat
    )
  )
  logLik(res)
}

tau <- seq(1, 2, length.out = 11)

# lik <- sapply(lambda, cond_lik)  # (sequential)
lik <- foreach(x = tau, .combine = c) %dopar% cond_lik(x)

ggplot(data.frame(tau, lik)[6:11,], aes(tau, lik)) + 
  geom_line()
```


## Recalculating the covariance for the hybrid-pure relationships by simulation

```{r sim-cov}

sample_bv <- function(sigma2) {
  ## Simulate two parents from E
  ## take the second, and simulate another parent from J
  x <- rnorm(2, sd = sqrt(sigma2['E']))
  y <- c(x[2], rnorm(1, sd = sqrt(sigma2['J'])))
  
  ## Compute BVs of half-sibs with mean-parents and MST
  xx <- mean(x) + rnorm(1, sd = sqrt(sigma2['E']/2))
  yy <- mean(y) + rnorm(1, sd = sqrt(sum(sigma2[c('E', 'J')])/4))
  
  return(c(xx, yy))
}

simbvs <- foreach(x = 1:1e4, .combine = 'rbind') %dopar% sample_bv(sigma2)
cov(simbvs)

```

The variance of the E is $\sigma^2_E = 9$, as expected.
The variance of the hybrid is $\frac{\sigma^2_E + \sigma^2_J}{2} = 7.5$, as also expected.
However, their covariance as half-sibs seems not to be $\frac14 \frac{3 \sigma^2_E + \sigma^2_J}{4} = 2.0625$ as was introduced in the model, but is rather close to $\frac14 \sigma^2_E = 2.25$.


## A new model for scaling the relationship matrix

```{r likelihood-profiling-the-other-way-around}

## Introduce the corresponding scaling factors 
## in the relationship matrix
rescale_A <- function(x) {

  A <- pedigreemm::getA(ped)
  
  ## The pure E subpopulations get multiplied by tau
  S <- A
  E.idx <- c(idx_pop('E'), idx_pop('EE'))
  S[E.idx, E.idx] <- A[E.idx, E.idx] * x
  
  ## The pure J subpopulations remains the same
  J.idx <- c(idx_pop('J'), idx_pop('JJ'))
  
  ## The hybrids related wuth pure E get a factor of tau
  S[idx_pop('EJ'), E.idx] <- A[idx_pop('EJ'), E.idx] * x
  S[E.idx, idx_pop('EJ')] <- A[E.idx, idx_pop('EJ')] * x
  
  ## The hybrids related wuth pure J remain the same
  S[idx_pop('EJ'), J.idx] <- A[idx_pop('EJ'), J.idx] 
  S[J.idx, idx_pop('EJ')] <- A[J.idx, idx_pop('EJ')] 

  ## Finally, the hybrids related with other hybrids get a factor of (1+tau)/2
  S[idx_pop('EJ'), idx_pop('EJ')] <- A[idx_pop('EJ'), idx_pop('EJ')] * (x+1)/2

  return(S)
}

S <- rescale_A(1.5)
idx <- -c(idx_pop('E'), idx_pop('J'))  # exclude founders
res <- remlf90(y ~ sp,
               generic = list(
                 J = list(incidence  = Z[, idx],
                          covariance = S[idx, idx])),
               data = dat
)

summary(res)
```

Still...
