---
title: "Handling pedigrees"
author: "Facundo Muñoz"
date: "`r paste(Sys.Date(), 'breedR version:', packageVersion('breedR'))`"
output:
  pdf_document:
    toc: true
    toc_depth: 1
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 1
vignette: >
  %\VignetteIndexEntry{Handling pedigrees}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, include=FALSE}
library(breedR)
```


## What is a _pedigree_ 

- A 3-column `data.frame` or `matrix` with the codes for each individual and its parents

- A **family** effect is easily translated into a pedigree:
    - use the **family code** as the identification of a fictitious **mother**
    - use `0` or `NA` as codes for the **unknown fathers**
    
- A pedigree sintetizes **any kind of (genetic) relationship** between individuals from one or more generations

```{r pedigree, purl=FALSE, results='asis', echo=FALSE}
knitr::kable(head(globulus[, 1:3]))
```


## Checking pedigrees 

- For computational reasons, the pedigree needs to meet certain conditions:
    - Completness: all the individuals (also parents) must have an entry
        - with possibly unknown parents (code `0` or `NA`)
    - The offspring must follow the parents 
    - The codes must be sorted increasingly
    - The codes must be consecutive

- So, not every 3-column `data.frame` or `matrix` with codes is a proper pedigree:

```{r check-pedigree}
set.seed(123); n.ped <- 5
ped.nightmare <- matrix(sample(30, n.ped*3), n.ped, 3,
                        dimnames = list(NULL, c('self', 'sire', 'dam')))
check_pedigree(ped.nightmare)
```


## Building pedigrees

- **breedR** implements a _pedigree constructor_ that completes, sorts and recodes as necessary

- The resulting object, of class `pedigree` is guranteed to meet the conditions

```{r build-pedigree}
ped.fix <- build_pedigree(1:3, data = ped.nightmare)
check_pedigree(ped.fix)
attr(ped.fix, 'map')  # map from old to new codes
```

```{r compare-pedigrees, results = 'asis', echo=FALSE}
knitr::kable(ped.nightmare)
knitr::kable(as.data.frame(ped.fix))
```


## Using a pedigree in an additive genetic effect

- just include your original pedigree information and let **breedR** fix it for you

```{r exercise-pedigree-dat}
test.dat <- data.frame(ped.nightmare, y = rnorm(n.ped))
res.raw <- remlf90(fixed   = y ~ 1,
                   genetic = list(model = 'add_animal',
                                  pedigree = ped.nightmare,
                                # pedigree = test.dat[, 1:3],  # same thing
                                  var.ini = 1,
                                  id = 'self'),
                   var.ini = list(resid = 1),
                   data    = test.dat)

## pedigree has been recoded!
length(ranef(res.raw)$genetic)

## The pedigree used in the model matches the one manually built
identical(ped.fix, get_pedigree(res.raw))
```


## Recovering Breeding Values in the original coding
```{r PBV, results = 'asis'}
## Predicted Breeding Valuess of the observed individuals
## Left-multiplying the vector of BLUP by the incidence matrix
## gives the BLUP of the observations in the right order.
Za <- model.matrix(res.raw)$random$genetic  # incidence matrix
gen.blup <- with(ranef(res.raw),
                 cbind(value=genetic,
                       's.e.'=attr(genetic, 'se')))
PBVs <- Za %*% gen.blup
rownames(PBVs) <- test.dat$self
```

```{r PBV-table, results = 'asis', echo=FALSE}
knitr::kable(PBVs, digits = 2)
```

