---
title: "Heterogeneous variances and weighting"
author: "Facundo Muñoz"
date: "`r paste(Sys.Date(), 'breedR version:', packageVersion('breedR'))`"
output:
  pdf_document:
    toc: true
    toc_depth: 1
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 1
vignette: >
  %\VignetteIndexEntry{Heterogeneous variances and weighting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, include=FALSE}
library(breedR)
library(ggplot2)
theme_set(theme_bw())
```

By default, the Linear Mixed Models fitted with breedR assume
*homoscedasticity*. Meaning that given all the fixed and random effects, the
unexplained variation follow a Normal distribution with residual variance
$\sigma^2$.

Mathematically, that $\varepsilon \sim \mathcal N (0, \mathbf I \sigma^2)$ in the model
equation $$ y = X \beta + Z u + \varepsilon $$

Sometimes this is obviously wrong, and we need models where some observations
are observed with more or less residual variability than others

Here are a few common situations where heterogeneous variances is needed:

- The observations are actually derived or calculated from real measurements, such
as an average. Thus, the variance depends on the number of averaged measurements
(e.g. Daughter Yield Deviation measures).

- The observations are spread in time, and you want to model the residual variance
as a function of time (e.g. longitudinal models).


# Using **weights**

If the relative variation in the residual variances is know or can be estimated,
it can be specified as a vector of *weights* $w$, such that
$$\varepsilon \sim \mathcal N (0, (w^{-1/2})'\mathbf I w^{-1/2} \sigma^2).$$
In other words, the residual variance for the observation $i$ is $\sigma^2/w_i$.

Here is a simulation example of how to specify weights.

```{r sim-data}
set.seed(123)

n <- 1e3   # n obs
sigma2 <- 4  # true residual variance (for a weight of 1)
w = runif(n, min = .5, max = 2)  # vector of weights

dat <- 
  transform(
    data.frame(
      e = rnorm(n, sd = sqrt(sigma2))
    ),
    y = 10 + e/sqrt(w)  # simulated phenotype
  )
```

```{r wheights-example}
res <- remlf90(
  y ~ 1,
  data = dat,
  weights = w  # specification of weights
)
```

Note that the estimated residual variance is close to the true value. On the
other hand, the residual prediction-error are expected to have non-constant
variance.

```{r weights-results}
summary(res)

ggplot(transform(dat, est_e = residuals(res)), aes(e, est_e)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "darkgray")
```


# Estimating residual variance heterogeneity

This is currently not available in breedR.

Different group-wise residual variances (e.g. multi-site) can be easily induced
by using group-specific random effects.

For the general case, here are some notes to allow for some manual hacking if
needed.

In general, we need to estimate a residual variance parameter as a function of
some other variable $x$.
We then write the residual variance as a linear combination of a few base functions:
$$\sigma^2(x) = \sum_{k=0}^K \psi_k(x)\, r_k = \mathbf{\Psi r},$$
where the parameters $\sigma_k$ are to be estimated.

This covers the case both for group-wise residual variances (such as multi-site,
using a categorical variable $x$) or a continuously varying residual variance.

For the first case, the variable $x$ is categorical, taking a finite number of
values $K$, and we define $\psi_k$ as the corresponding indicator functions.

For the continuous case, the variable $x$ is continuous (e.g. age, temperature)
and the base functions can be Legendre polynomials, splines, etc. up to some 
arbitrary order $K$.

We need to manually build the matrix $\mathbf\Psi$, and exploit the PROGSF90
options `hetres_pos` and `hetres_pol` (see [documentation](http://nce.ads.uga.edu/wiki/doku.php?id=readme.aireml)).
