---
title: "Missing values"
author: "Facundo Muñoz"
date: "`r paste(Sys.Date(), 'breedR version:', packageVersion('breedR'))`"
output:
  pdf_document:
    toc: true
    toc_depth: 1
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 1
vignette: >
  %\VignetteIndexEntry{Missing values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


The handling of missing values (i.e. `NA`) depends on *where* they are.

## Missing response

It is perfectly valid to have missing vaules in the dependent variable. There is
no need of removing those individuals from the dataset. Furthermore, including
them will yield
[predictions](https://github.com/famuvie/breedR/wiki/Overview#prediction) for
their phenotype, based on the predictive variables.

```{r missing-response, message=FALSE}
library(breedR)

N <- 1e3
x <- rep(1:4, each = N/4)
dat <- data.frame(y = x + rnorm(N),
                  x = factor(letters[x]))
dat$y[1] <- NA
head(dat)
res <- remlf90(y ~ x, data = dat)

## The predicted phenotype for y[1] is the estimated effect
## of the corresponding level of x
fitted(res)[1] == fixef(res)$x['a']
```



## Missing value for a fixed effect

This is not allowed, as it would yield an underdetermined system of equations.
`breedR` issues an error if missing values are detected.

```{r missing-fixed, message = FALSE, error = TRUE}
N <- 1e3
x <- rep(1:4, each = N/4)
dat <- data.frame(y = x + rnorm(N),
                  x = factor(letters[x]))
dat$x[c(1, 3, 5)] <- NA
head(dat)
res <- remlf90(y ~ x, data = dat)
```

Idem for a regression variable.

```{r missing-fixed-regression, message = FALSE, error = TRUE}
N <- 1e3
x <- runif(N)
dat <- data.frame(y = 1 + 2*x + rnorm(N),
                  x = x)
dat$x[c(1, 3, 5)] <- NA
head(dat)
res <- remlf90(y ~ x, data = dat)
```



## Missing value for a random effect

These **are** allowed.
The incidence matrix will have a row of zeros for the corresponding individual.

```{r missing-diagonal, message = FALSE}
N <- 1e3
N.blk <- 20
blk.effects <- rnorm(N.blk, sd = 2)
blk.idx <- sample(seq_len(N.blk), N, replace = TRUE)
dat <- data.frame(y = 1 + blk.effects[blk.idx] + rnorm(N),
                  blk = factor(blk.idx))
dat$blk[1] <- NA
head(dat)

res <- remlf90(y ~ 1, random = ~ blk, data = dat)

sum(model.matrix(res)$blk[1,])
```

As a consequence, the predicted phenotype will be based on the remaining
available effects. In this case, the global mean.


```{r missing-diagonal-residual}
fitted(res)[1] == fixef(res)$Intercept[1]
```


The spatial block effect is another way of writing the previous experiment.
So it works in the same way.

```{r missing-block, message = FALSE}
coord <- expand.grid(row = 1:20, col = 1:50)
res <- remlf90(y ~ 1,
               spatial = list(model = 'blocks',
                              coord = coord,
                              id    = 'blk'), 
               data = dat)

c(sum(model.matrix(res)$spatial[1,]) == 0,
fitted(res)[1] == fixef(res)$Intercept[1])
```



However, the empirical residuals of the individuals with missing values of the
random effects will have an increased variance.
We can show that by replicating the previous experiment and computing the
variance of the residual for the first observation.

```{r sample-residual-function, echo = FALSE}
sample_first_residual <- function(N = 1e3, N.blk = 20) {
  blk.effects <- rnorm(N.blk, sd = 2)
  blk.idx <- sample(seq_len(N.blk), N, replace = TRUE)
  dat <- data.frame(y = 1 + blk.effects[blk.idx] + rnorm(N),
                    blk = factor(blk.idx))
  dat$blk[1] <- NA
  res <- suppressMessages(remlf90(y ~ 1, random = ~ blk, data = dat))
  return(unname(resid(res)[1]))
}
```


```{r variance-missing-residuals}
resid_sample <- replicate(1e2, sample_first_residual())
var(resid_sample)

```

This can be important when fitting several random effects.
See below.


## Missing values in genetic effects

For an additive genetic effect, the relationship between individuals is given in
the pedigree.
It is legitimate not knowing the relatives for some individual.
This is what happens with founders, for example.

Use `NA` for unknown relatives.
If both are unknown (e.g. founders), the genetic effect (Breeding Value) will be
predicted based on its phenotype, the other effects, and the estimated
heritability.


```{r missing-add_animal, message = FALSE}
dat <- breedR.sample.phenotype(
  fixed = c(mu = 10, x = 2),
  genetic = list(model    = 'add_animal',
                 Nparents = c(10, 10),
                 sigma2_a = 2,
                 check.factorial = FALSE),
  N = 1e3)
head(dat)

res <- remlf90(phenotype ~ 1 + X.x,
               genetic = list(model = 'add_animal',
                              pedigree = dat[, 1:3],
                              id    = 'self'),
               data = dat)

str(ranef(res)$genetic)
```



**Important issue**
Having random effects with missing values in **combination** with genetic
models, can yield spurious predictions of Breeding Values.
This is due to the higher variability of the residual term, for the individuals
with missing values in random effects.


## Missing values in coordinates of spatial effects

Are allowed. 
Just like in any other random effect.
For those cases, the spatial component will not participate in the prediction.

```{r missing-coordinates, message = FALSE}
dat <- breedR.sample.phenotype(
  fixed = c(mu = 10, x = 2),
  spatial = list(model     = 'AR',
                 grid.size = c(10, 5),
                 rho       = c(.2, .8),
                 sigma2_s  = 1)
)
dat$Var1[1] <- NA
head(dat)

res <- remlf90(phenotype ~ 1 + X.x,
               spatial = list(model = 'AR',
                              coord = dat[, c('Var1', 'Var2')],
                              rho   = c(0.2, 0.8)),
               data = dat)

sum(model.matrix(res)$spatial[1,])
```

